Binning Strategies and Related Loss for Large Data

### 2015-03-30 Paper.Rnw
more work on section 5.1

\subsection{Binning Specification and Spatial Loss}

To explore the spatial loss properties under different binning approaches we begin with a comparison of standard and random binning. If equivalent bin widths are used, the net spatial loss under standard binning will always be less than or equal to the net spatial loss under random binning. This is because the minimal spatial loss for each data point under random binning is to allocate to the nearest bin center, which is how the point would be allocated in standard binning.

Figure~\ref{fig:SpatialLossCompare} displays the proportion of potential spatial loss from binning each set of simulated data with standard and random binning algorithms using square bins ranging from 2 units to 20 units. It is clear to see that the spatial loss for random binning is always higher than for standard binning and that the loss grows as binwidth increases. Also, the proportion of potential spatial loss from random binning increases with binwidth at a faster rate than with standard binning, as indicated by the widening gap and steeper slopes.

@@

>>>>

\subsection{Rectangular Binning Specifications and Spatial Loss}

For rectangular binning there are many specification options; including type of algorithm, location of the origin and binwidths for each dimension. To explore the spatial loss properties under different binning approaches we begin with a comparison of standard and random binning. If equivalent bin widths are used, the net spatial loss under standard binning will always be less than or equal to the net spatial loss under random binning. This is because the minimal spatial loss for each data point under random binning is to allocate to the nearest bin center, which is how the point would be allocated in standard binning.

Figure~\ref{fig:SpatialLossCompare} displays the proportion of potential spatial loss from binning each set of simulated data with standard and random binning algorithms using square bins with a sizes ranging from 2 units$^2$ to 20 units$^2$. It is clear to see that the spatial loss for random binning is always higher than for standard binning and that the loss grows as binwidth increases. The proportion of potential spatial loss from random binning increases with binwidth at a faster rate than with standard binning, as indicated by the widening gap and steeper slopes.

@@

All rectangular binning algorithms require the specification of a bin width for each dimension and a binning origin. Smaller binwidths coincide with smaller spatial losses, but reducing bin sizes comes with a few important caveats. With smaller bins comes more computation requirements; a non-negligable consideration in settings with truly massive data. Additionally, the binwidth should be an integer multiple of the resolution of the data to avoid undesirable visual attributes in the binned scatterplot.

Consider the simulated uniform data, we have the fine version and the coarse version, which rounded to nearest even integer. Figure~\ref{fig:strippedplots} displays the binned scatterplots under several scenarios.
